---
title: "Data Analysis"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls())
```

```{r, include=FALSE}
library(isdas)
library(spatstat)
library(tidyverse)
library(ggplot2)
library(rgdal)
library(sf)
library(dplyr)
library(plotly)
library(webshot)
library(gridExtra)
library(patchwork)
library(spdep)
```

# Data Processing

```{r}
# Set working directory to the project folder on github
wd <- getwd()

#import shp file of neighbourhoods
nshp <- st_read(paste(wd, "/Neighbourhoods.shp", sep = ""))
st_drivers()

#visualizing shpfile
ggplot() + 
  geom_sf(data = nshp, size = 1, color = "black", fill = "red") + 
  ggtitle("Neighbourhoods") + 
  coord_sf()
```

```{r}
#Deleting the neighbourhood ID from the neighbourhood field for pairing to final table

#Extracting column as matrix
f2 <- as.matrix(nshp[2])

#deleting geometry field
f2 <- subset(f2, select = -c(geometry) )

#deleting neighbourhood IDs
f2 <- gsub("[1234567890()]", " ", f2)

#deleting space after
f2 <- str_squish(f2)

#rejoining to shp file
nshp[2] <- f2
```

```{r}
#Covid Cases
covidcases <- read_csv(paste(wd,"/COVID19 cases.csv", sep=""))
#Neighborhood Data 
neighbourhood <- read_csv(paste(wd,"/Neighbourhood_POP.csv", sep=""))
```

```{r}
chr <- as.character.Date(covidcases$`Episode Date`)
              
covidcases$'nd' <-chr 

fweek <- c("2021-12-01","2021-12-02","2021-12-03","2021-12-04","2021-12-05","2021-12-06","2021-12-07")
lweek <- c("2021-12-25","2021-12-26","2021-12-27","2021-12-28","2021-12-29","2021-12-30","2021-12-31")

fweektb <- subset(covidcases,nd %in% fweek)
lweektb <- subset(covidcases,nd %in% lweek)

ftab <- as.data.frame(table(fweektb$`Neighbourhood Name`)) 
ltab <- as.data.frame(table(lweektb$`Neighbourhood Name`)) 
```

```{r}
covidcases$`Episode Date`[1]
```

```{r}
#This section in creating final_table 

#Neighbourhood_Name, F_weekcases, L_weekcases, POP, F_weekrate,, L_weekrate, Rate_over_month, Case_over_month

#We need to add population for each neighbourhood and join it to the final table 

Neighbourhood <- ftab$Var1
F_weekcases <- ftab$Freq
L_weekcases <- ltab$Freq
Case_over_month <- ltab$Freq - ftab$Freq

#Create Final table 
Final_Table <- data.frame(Neighbourhood, F_weekcases, L_weekcases)

```

```{r}
#3 Neighbourhood are different so there must be an manual correction prior to merging 

neighbourhood$Neighbourhood[128] <- "Weston-Pellam Park"
neighbourhood$Neighbourhood[16] <- "Briar Hill - Belgravia"
neighbourhood$Neighbourhood[30] <- "Danforth-East York"

```

```{r}
#Merge neighbour with final table
Final_Table <- merge(Final_Table, neighbourhood)

# Creating the case rates per 10,000  
Final_Table$F_weekrate <- Final_Table$F_weekcases/Final_Table$POP * 10000
Final_Table$L_weekrate <- Final_Table$L_weekcases/Final_Table$POP * 10000

# Creating rate change 
Final_Table$Rate_Change <- Final_Table$L_weekrate - Final_Table$F_weekrate

# Creating fold increase 
Final_Table$fold_increase <- Final_Table$L_weekrate/Final_Table$F_weekrate 
```

```{r}
#Neighbourhood Income Data
Socioeconomic_variables <- read_csv(paste(wd,"/Neighbourhoods_socioeconomic_variables.csv", sep=""))

```
```{r}
#3 neighbourhoods are different so there must be an manual correction prior to merging 
Socioeconomic_variables$Neighbourhood[128] <- "Weston-Pellam Park"
Socioeconomic_variables$Neighbourhood[16] <- "Briar Hill - Belgravia"
Socioeconomic_variables$Neighbourhood[30] <- "Danforth-East York"
```

```{r}
#Merge neighbourhood socioeconomic variables data with final table
Final_Table <- merge(Final_Table, Socioeconomic_variables)
```

```{r}
#Merging shpfile to final table

nshp$FIELD_2[22] <- "Danforth-East York"
nshp$FIELD_2[25] <- "North St. James Town"
nshp$FIELD_2[93] <- "Mimico (includes Humber Bay Shores)"
nshp$FIELD_2[114] <- "Briar Hill - Belgravia"
nshp$FIELD_2[115] <- "Cabbagetown-South St. James Town"

nshp <- nshp %>%
  rename(Neighbourhood = FIELD_2)
Final_Table <- left_join(nshp, Final_Table, by = "Neighbourhood")

Final_Table <- Final_Table %>%
  rename(`_id` = FIELD_1,
          AREA_ATTR_ID = FIELD_3,
          PARENT_AREA_ID = FIELD_4,
          AREA_SHORT_CODE = FIELD_5,
          AREA_LONG_CODE = FIELD_6,
          AREA_NAME = FIELD_7,
          AREA_DESC = FIELD_8,
          X = FIELD_9,
          Y = FIELD_10,
          LONGITUDE = FIELD_11,
          LATITUDE = FIELD_12,
          OBJECTID = FIELD_13,
          Shape__Area = FIELD_14,
          Shape__Length = FIELD_15)

```

# Data Analysis

## Initial Exploration/Visualization

### Dependent Variable
```{r}
# have change in COVID case rates (cases per 10000) between first and last week
# by difference and fold-change

ggplot(data = Final_Table) +
  geom_sf(aes(fill = fold_increase)) + 
  scale_fill_distiller(name = "Fold Change",
                       palette = "YlOrRd", 
                       direction = 1) +
  ggtitle("Changes in COVID-19 Rates (Cases per 10,000) from... \n to... in Toronto by Neighbourhood") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```
```{r}
ggplot(data = Final_Table) +
  geom_sf(aes(fill = F_weekrate)) + 
  scale_fill_distiller(name = "Cases Per 10000",
                       palette = "YlOrRd", 
                       direction = 1) +
  ggtitle("COVID-19 Cases in Toronto by Neighbourhood \n During the First Week of December 2021") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```
```{r}
ggplot(data = Final_Table) +
  geom_sf(aes(fill = L_weekrate)) + 
  scale_fill_distiller(name = "Cases Per 10000",
                       palette = "YlOrRd", 
                       direction = 1) +
  ggtitle("COVID-19 Cases in Toronto by Neighbourhood \n During the Last Week of December 2021") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# Final_Table$relative_change <- (Final_Table$L_weekrate - Final_Table$F_weekrate) /Final_Table$F_weekrate 
# 
# ggplot(data = Final_Table) +
#   geom_sf(aes(fill = relative_change)) + 
#   scale_fill_distiller(name = "Relative Change",
#                        palette = "YlOrRd", 
#                        direction = 1) +
#   theme_minimal()

# fold and relative change have similar values
```

### Independent Variables(s)
```{r}
# have average income and proportion of 18-24 y/os
# population density can be calculated w/ POP and Shape_Area too

# rename column names so it's easier to call
Final_Table <- Final_Table %>%
  rename(avg_income = `Average income $ (Total income in 2010 of population aged 15 years and over)`,
         perc_18_24 = `Percent of Pop Age 18 to 24`)

ggplot(data = Final_Table) +
  geom_sf(aes(fill = avg_income)) + 
  scale_fill_distiller(name = "Average Income ($)",
                       palette = "YlOrRd", 
                       direction = 1) +
  ggtitle("Average Income in Toronto by Neighbourhood") #maybe adjust this later for detail
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```
```{r}
ggplot(data = Final_Table) +
  geom_sf(aes(fill = perc_18_24)) + 
  scale_fill_distiller(name = "Percentage of Population",
                       palette = "YlOrRd", 
                       direction = 1) +
  ggtitle("Percentage of People Aged 18-24 in Toronto by Neighbourhood") #maybe adjust this later for detail (e.g., as of...)
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# if we want to try population density

Final_Table$pop_density <- Final_Table$POP / Final_Table$Shape__Area

ggplot(data = Final_Table) +
  geom_sf(aes(fill = pop_density)) + 
  scale_fill_distiller(name = "Population Density",
                       palette = "YlOrRd", 
                       direction = 1) +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
```
## Testing Dependent Variable for Randomness

Q: is the pattern of changes in COVID-19 rates random or non-random?

### Spatial Moving Averages
```{r}
Final_Table.nb <- poly2nb(pl = Final_Table)
Final_Table.w <- nb2listw(Final_Table.nb)

fold_increase.sma <- lag.listw(Final_Table.w, Final_Table$fold_increase)
Final_Table$fold_increase.sma <- fold_increase.sma

# null landscapes
fold_null1 <- sample(Final_Table$fold_increase)
fold_null1.sma <- lag.listw(Final_Table.w, fold_null1)
Final_Table$fold_null1 <- fold_null1
Final_Table$fold_null1.sma <- fold_null1.sma

fold_null2 <- sample(Final_Table$fold_increase)
fold_null2.sma <- lag.listw(Final_Table.w, fold_null2)
Final_Table$fold_null2 <- fold_null2
Final_Table$fold_null2.sma <- fold_null2.sma

fold_null3 <- sample(Final_Table$fold_increase)
fold_null3.sma <- lag.listw(Final_Table.w, fold_null3)
Final_Table$fold_null3 <- fold_null3
Final_Table$fold_null3.sma <- fold_null3.sma

fold_null4 <- sample(Final_Table$fold_increase)
fold_null4.sma <- lag.listw(Final_Table.w, fold_null4)
Final_Table$fold_null4 <- fold_null4
Final_Table$fold_null4.sma <- fold_null4.sma

fold_null5 <- sample(Final_Table$fold_increase)
fold_null5.sma <- lag.listw(Final_Table.w, fold_null5)
Final_Table$fold_null5 <- fold_null5
Final_Table$fold_null5.sma <- fold_null5.sma
```

```{r}
Final_Table2 <- Final_Table %>% 
  select(fold_increase.sma, 
         fold_null1.sma,
         fold_null2.sma,
         fold_null3.sma,
         fold_null4.sma,
         fold_null5.sma,
         geometry) %>%
  gather(VAR, fold_sma, -geometry)

Final_Table2 <- Final_Table2 %>%
  data.frame(Final_Table %>%
               st_drop_geometry() %>%
               select(fold_increase,
                      fold_null1,
                      fold_null2,
                      fold_null3,
                      fold_null4,
                      fold_null5,) %>% 
               gather(VAR, fold) %>% 
               select(fold))
```

```{r}
ggplot(data = Final_Table2, aes(x = fold, y = fold_sma, color = VAR)) +
  geom_point(alpha = 0.1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_smooth(method = "lm") +
  coord_equal() +
  facet_wrap(~ VAR, ncol = 2)
```


### Moran's I
```{r}
mp <- moran.plot(Final_Table$fold_increase, Final_Table.w)
moran.test(Final_Table$fold_increase, Final_Table.w)
```
### Local Moran's I
```{r}
fold_increase.local <- localmoran(Final_Table$fold_increase, Final_Table.w)
colnames(fold_increase.local) <- c("Ii", "E.Ii", "Var.Ii", "Z.Ii", "p.val")

Final_Table <- Final_Table %>% 
  left_join(data.frame(Neighbourhood = Final_Table$Neighbourhood, 
                       fold_increase.local), 
            by = "Neighbourhood")
```

```{r}
# plot_ly(Final_Table) %>%
#     add_sf(type = "scatter",
#            split = ~(p.val < 0.05), 
#            color = ~Type, 
#            colors = c("red", 
#                       "khaki1", 
#                       "dodgerblue", 
#                       "dodgerblue4")) 
```



## Regression Analysis

Q: what may be making the pattern of changes in COVID-19 rates non-random?

### Model 1: Linear Regression w/ Income
```{r}
model1 <- lm(formula = fold_increase ~ avg_income, data = Final_Table)
summary(model1) 
```

```{r}
ggplot(data = Final_Table, aes(x = avg_income, 
                               y = fold_increase)) + 
  geom_point() +
  geom_abline(slope = model1$coefficients[2], 
              intercept = model1$coefficients[1], 
              color = "blue", size = 1) +
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0)
```

```{r}
# analyzing the residuals

Final_Table$model1.e <- model1$residuals

plot_ly(Final_Table) %>% 
  add_sf(type = "scatter",
         color = ~(model1.e > 0), 
         colors = c("red", 
                    "dodgerblue4")) 
```

```{r}
moran.test(Final_Table$model1.e, 
           Final_Table.w)
```

### Model 2: Inverse Income

```{r}
Final_Table <- mutate(Final_Table, 
                      inv_income = 1/avg_income)

model2 <- lm(formula = fold_increase ~ inv_income, 
             data = Final_Table)
summary(model2) 
```

```{r}
ggplot(data = Final_Table, 
       aes(x = avg_income, 
           y = fold_increase)) + 
  geom_point() +
  stat_function(fun=function(x)model2$coefficients[1] + model2$coefficients[2]/x, 
                geom="line", 
                color = "blue",
                size = 1) +
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0)
```
```{r}
Final_Table$model2.e <- model2$residuals

moran.test(Final_Table$model2.e, 
           Final_Table.w)
```

### Model 3: Log Fold Increase
```{r}
Final_Table <- Final_Table %>%
  mutate(ln_fold = log(fold_increase)) 

model3 <- lm(formula = ln_fold ~ avg_income, 
             data = Final_Table)

summary(model3)
```
```{r}
ggplot(data = Final_Table, 
       aes(x = avg_income, 
           y = fold_increase)) + 
  geom_point() +
  stat_function(fun=function(x)exp(model3$coefficients[1] + model3$coefficients[2] * x), 
                geom="line", 
                color = "blue", 
                size = 1) +
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0)
```

```{r}
Final_Table$model3.e <- model3$residuals

moran.test(Final_Table$model3.e, 
           Final_Table.w)
```


### Model 4: Linear Regression with Income + 18-24s + Density
```{r}
model4 <- lm(formula = fold_increase ~ avg_income + perc_18_24 + pop_density, data = Final_Table)
summary(model4) 
```

### Model 5: Quadratic Income + Quadratic Density + 18-24 y/os

```{r}
Final_Table <- Final_Table %>%
  mutate(quad_income = avg_income**2, 
         quad_density = pop_density**2)

model5 <- lm(formula = fold_increase ~ quad_income + perc_18_24 + quad_density, data = Final_Table)
summary(model5) 
```

## Model 6: Quadratic Income Only
```{r}
model6 <- lm(formula = fold_increase ~ quad_income, data = Final_Table)
summary(model6) 

ggplot(data = Final_Table, 
       aes(x = avg_income, 
           y = fold_increase)) + 
  geom_point() +
  stat_function(fun=function(x)model6$coefficients[1] + model6$coefficients[2] * x**2, 
                geom="line", 
                color = "blue", 
                size = 1) +
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0)

Final_Table$model6.e <- model6$residuals

moran.test(Final_Table$model6.e, 
           Final_Table.w)
```

### Model 7: Inverse Income, Log Fold Increase
```{r}
model7 <- lm(formula = ln_fold ~ inv_income, data = Final_Table)
summary(model7) 

ggplot(data = Final_Table, 
       aes(x = avg_income, 
           y = fold_increase)) + 
  geom_point() +
  stat_function(fun=function(x)exp(model7$coefficients[1] + model7$coefficients[2]/x), 
                geom="line", 
                color = "blue", 
                size = 1) +
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0) +
  ylim(c(0,300))

Final_Table$model7.e <- model7$residuals

moran.test(Final_Table$model7.e, 
           Final_Table.w)
```


