---
title: "Income as a Determinant of the Speed of Transmission: Study of the Fifth Wave of SARS-CoV-2 in Toronto, Ontario"
#subtitle: "A subtitle"
thanks: ""
author:
- name: Jason Lam
  student_number: 400201299
- name: Emily Fletcher
  student_number: 400210334
- name: Syed Hammad Uddin
  student_number: 400188361
- name: Jenny Chau
  student_number: 400233137
- name: Pierce Bourgeois 
  student_number: 400220858
subject: "ENVSOCTY 4GA3"
abstract: ""
keywords: "Income, Toronto, COVID-19, Spatial Statisitics, Spatial Analysis"
date: "4/22/2022"
output:
  pdf_document:
    # The project-template-default.tex file was minimally adapted from Steven V. Miller's template for academic manuscripts. See:
    # http://svmiller.com/blog/2016/02/svm-r-markdown-manuscript/
    # https://github.com/svmiller/svm-r-markdown-templates/blob/master/svm-latex-ms.tex
    template: project-template-default.tex
bibliography: [bibliography.bib, packages.bib]
always_allow_html: true
---

<!--Chunks of code can have names; the chunk option "include" controls whether the chunk and its output are printed in the final document-->
```{r load-packages, include=FALSE}
library(isdas)
library(spatstat)
library(tidyverse)
library(ggplot2)
library(rgdal)
library(sf)
library(dplyr)
library(gridExtra)
library(patchwork)
library(spdep)
```

```{r write-package-bib, include=FALSE}
# This function is used to write a bibliography for the `R` packages used in the paper
knitr::write_bib(file = 'packages.bib')
```

```{r read-census-data, include=FALSE}
```

<!--This create a page break, i.e., starts a new page-->
<!--\newpage-->

<!-- 
To cite references in your bibliography.bib file, use [@item] if you want it to be cited in brackets, or @item if you want it to be cited as Name (year). If you want to cite various items in brackets, separate them with semicolons [@item1; @item2]
-->


# Introduction
Since the first cases in 2019, the COVID-19 outbreak has taken the world by storm. As of writing this (April 2022), there are over five hundred million cases reported worldwide [@WHO_2022]. With its ability to rapidly spread, the COVID-19 has had profound consequences in every facet of public and private life. Among these include medical, economic, and social spaces [@Baena-Diez_Barroso_Cordeiro-Coelho_Diaz_Grau_2020].  Most importantly, these unsettling times have shone a light on pre-existing disparities within national health systems. For many researchers, this does not come as a surprise as socioeconomic factors influencing health outcomes have been noted ubiquitously throughout the world [@Khalatbari-Soltani_Cumming_Delpierre_Kelly-Irving_2020]. According to Khalatbari-Soltani et al. (2020), there is even a link between those socio-economically disadvantaged and increased risk of infectious disease.  

This research continues in a vein of spatial statistical studies which analyze COVID-19 incidence rates paired with a socio-economic factor. Building upon Khalatabari-Soltani, et al. (2020) and the socio-economic position (SEP) framework, this study seeks to examine the relationship between income levels, population density, and proportion of young adults against COVID-19 incidence rates. More specifically, the overarching objective of this study is to analyze the relationship between the noted socio-economic variables and incidence rates of COVID-19 during the fifth wave in Toronto, Ontario, Canada.

  
# Background

A person’s income level affects how they go about their everyday lives. The associated behavioural pattern has been linked to change in health outcomes [@Khalatbari-Soltani_Cumming_Delpierre_Kelly-Irving_2020].  Low income, for example, affects housing condition and leads to more tight housing arrangements. Such factors have been associated in the increased risk of infections for pathogens such as tuberculosis [@Khalatbari-Soltani_Cumming_Delpierre_Kelly-Irving_2020].  

In Ontario, from January 21 to June 30, 2022, the most attributed workplace was manufacturing [@Murti_Achonu_Smith_Brown_Kim_Johnson_Ravindran_Buchan_2021]. Manufacturing accounted for 45% of outbreaks which totaled 65% of outbreak cases. Another notable sector was Transport and Warehousing (11% of outbreaks, 8% of outbreak cases).  In Toronto, it has been observed the COVID-19 first infiltrate in high income communities before quickly spreading to lower income communities [@Mishra_Ma_Moloney_Yiu_Darvin_Landsman_Kwong_Calzavara_Straus_Chan_et_al_2022].  According to Mishra et al. (2022) lower income neighbourhoods were also defined by their higher dwelling densities and greater proportion of occupations that could not make the transition to remote work.   

In addition, the susceptibility of adolescents (aged 10-19 years) and youth (aged 15-24 years) to COVID-19 has been a controversial research topic since the pandemic began [@Rumain_Schneiderman_Geliebter_2021]. While several studies have concluded that young adults are significantly less susceptible to COVID-19 than older adults, others have found that the prevalence of COVID-19 for adolescents and youth to be significantly greater than that of older adults [@Rumain_Schneiderman_Geliebter_2021]. In April 2021, COVID-19 cases were rising rapidly for young Canadians, with cases being highest among those aged 20 to 39 [@Aziz_2021].  Suggested factors that attribute to higher COVID-19 incidence among younger people include the reopening of high schools, colleges, and universities, larger and more frequent social gatherings and non-compliance with public health guidelines due to perceived low-risk of severe symptoms for the age group, and low income.  [@Aleta_Moreno_2020]. Health-related behaviours of younger adults may also affect their susceptibility to COVID-19 infection [@Abbasi_2020]. In an online national survey of adolescents and young adults, vaping and the dual use of e-cigarettes and cigarettes heavily increased the chances of COVID-19 diagnosis [@Gaiha_Cheng_Halpern-Felsher_2020].  

The first case of COVID-19 in Ontario (and Canada) was reported on January 25, 2020 [@Nielsen_2020]. As the virus began to spread, Ontario entered its first wave of COVID-19 on February 26, 2020. The first wave of COVID-19 lasted 188 days, ending on August 31, 2020 [@PHO_2021]. As Ontario began loosening restrictions as part of its 3-stage reopening plan, people started getting together again, and observed cases began to rise. Ontario’s second wave began September 2020 and ended in February 2021, with cases peaking in January 2021 [@PHO_2021]. The third wave in Ontario was driven by the Alpha (B.1.1.7) variant, which was more transmissible [@Detsky_Bogoch_2021]. The third wave lasted from March to July 2021, and was the largest wave yet, with the peak number of new cases in a day in Ontario being 5067 [@PHO_2022]. The emergence of the Delta variant (B.1.617.2) caused a smaller and shorter fourth wave in Ontario that lasted from August to October 2021. The largest number of new cases reported in a day in Ontario during the fourth wave was 878. 

```{r, echo=FALSE}
tab_01 = data.frame(
  Wave = c("1st", "2nd", "3rd", "4th", "5th"),
  Associated_Variant = c("Original Strain", "Original Strain", "Alpha", "Delta", "Omicron"),
  Approx_Start = c("February 2020 ", "September 2020 ", "March 2021 ", "August 2021 ", "December 2022 "),
  Approx_End = c("August 2020", "February 2021", "July 2021", "October 2021", "February 2022"),
  Peak = c("752", "4,168", "5,067", "878", "19,373"),
    Total = c("42,486", "260,643", "24,7654", "49,704", "469,955")
)
```

```{r, echo=FALSE}
tab_01 %>% knitr::kable(caption = "Waves of COVID-19 in Ontario", col.names = c("Wave", "Associated Variant", "Approx. Start", "Approx. End", "Peak Cases Per Day", "Total Cases"))
```

The fifth wave of the pandemic lasted from the beginning of December 2021 until mid-February 2022. The catalyst for this was the emergence of a new, highly transmissible variant called Omicron. The variant, which was first reported globally in November 2021, has been thoroughly researched due to its scale and rate of infection. This research suggests that the variant is highly transmissible due to several factors. This includes the fact that Omicron is more likely to evade immunity from a previous infection, meaning that there is a high chance that you can get re-infected with COVID-19 [@Pulliam_Schalkwyk_Govender_Gottberg_Cohen_Groome_Dushoff_Mlisana_Moultrie_2021]. Other research suggests that the variant is up to 3.7% more infectious among vaccinated citizens than its predecessors [@Mohsin_Mahmud_2022]. During the fifth wave, it became the dominant strain and was responsible for 95% infections globally. In Ontario, the first Omicron cases were reported on November 28, 2021 [@GofO_2021]. During the Omicron wave, the highest number of new cases reported for a single day in the province was 19,373 [@PHO_2022].  

Lastly, Toronto has the densest urban core in the province and is one of the most densely populated regions in North America. This has made it susceptible to the ability of COVID-19 to rapidly spread. To date, there have been more than 300,000 reported cases with more than 4000 deaths [@Toronto_2021]. Within the city, there are several pockets that are denser than others, and this density is an important factor to look at. Population density is a measure of spatial distribution of people across space. In the case of Toronto, St James Town is the most densely populated neighbourhood in the city [@CUI_2016]. Research around population density and its link to COVID-19 susceptibility is limited. Past literature has not shown a clear relationship between the two, with some noting a positive correlation [@Hamidi_Hamidi_2021] while others deducing an insignificant relationship [Carozzi_Provenzano_Roth_2020]. This, as suggested by the entire catalogue of research, is connected to the regional variations connected to density. Some denser areas may have better services to limit their exposure to the virus, while others may be poorer and so may be more susceptible to the virus. It is important to explore this phenomenon in the context of Toronto, to understand the type of relationship found in the city.  

# Study area 

The analysis was conducted at the neighbourhood level for the City of Toronto (See Figure 1). 'Neighbourhood' is a geographic level specifically designed by the City of Toronto. They were created by city to help government and other planning organizations with obtaining socio-economic data [@Toronto_2017]. In total there are 140 unique areas, and their boundaries are based on the Canadian Census Track. Each neighbourhood may contain between two to five of these census tracts. The geography of neighbourhood was chosen for this study, as opposed to wards or dissemination blocks, not only to showcase acute changes within populations but due to the availability of both COVID-19 and the socio-economic data. The large number neighbourhoods also enable this study to capture diversity across the city.


```{r, include=FALSE}
# Set working directory to the project folder on github
wd <- getwd()

#import shp file of neighbourhoods
nshp <- st_read(paste(wd, "/data/Neighbourhoods.shp", sep = ""))
```

```{r, echo=FALSE, fig.cap = "Neighbourhoods in Toronto, Ontario, Canada"}
#visualizing shpfile
ggplot() + 
  geom_sf(data = nshp, size = 0.6, color = "white", fill = "grey") + 
  coord_sf(datum = NA) + 
  theme_void()
```

# Data 

The COVID-19 data was retrieved from the City of Toronto Open Data portal [@TPH_2022]. The data was downloaded as a comma-separated values (CSV) file. This data is updated weekly by the city and reports each individual case as a record. The time period of interest was December 2021, corresponding to the fifth wave of the pandemic in Ontario. The cases from the first week of December 2021 (Dec. 1 - Dec. 7) and the last week (Dec. 25 - Dec. 31) were filtered out and aggregated by neighbourhood. The socio-economic data used in this study was also retrieved from Toronto Open Data and comes from a neighbourhood profile dataset [@TSDFA_2011].


# Methods

This study used RStudio to conduct both the data pre-processing, that is cleaning of the original datasets, as well as the analysis. 

The analysis was done using several R packages, including spatstat, tidyverse, ggplot2, dplyr, webshot, gridExtra, patchwork, and spdep to analyze the area data and find a potential relationship between the chosen independent variables () and the dependent variable () for the city of Toronto. For the initial visualization and analysis of the area data, choropleth maps and Moran’s I were used to identify patterns in the data. Then, to gain more insight on the process behind these patterns, regression analysis was used to determine the relationship between the independent and dependent variables.

A unique method was ran to covered the date data in the COVID-19 case table into character data type. This was so the records could be parsed in respect to this studies period of interest. 

This document was also written and exported through R-Markdown with minimal adaptation from Steven V. Miller's template for academic manuscripts. See: [http://svmiller.com/blog/2016/02/svm-r-markdown-manuscript/](http://svmiller.com/blog/2016/02/svm-r-markdown-manuscript/)[https://github.com/svmiller/svm-r-markdown-templates/blob/master/svm-latex-ms.tex] and (https://github.com/svmiller/svm-r-markdown-templates/blob/master/svm-latex-ms.tex)

The data and code used in this project can be found on Github: [https://github.com/lamj54/4GA3-Project](https://github.com/lamj54/4GA3-Project). This document is also available as a RMarkdown file, which includes both the code and the text used to create this report. 

<!--Data cleaning section-->
```{r, include = FALSE}
#Deleting the neighbourhood ID from the neighbourhood field for pairing to final table

#Extracting column as matrix
f2 <- as.matrix(nshp[2])

#deleting geometry field
f2 <- subset(f2, select = -c(geometry) )

#deleting neighbourhood IDs
f2 <- gsub("[1234567890()]", " ", f2)

#deleting space after
f2 <- str_squish(f2)

#rejoining to shp file
nshp[2] <- f2
```

```{r, include = FALSE}
#Covid Cases
covidcases <- read_csv(paste(wd,"/data/COVID19 cases.csv", sep=""))
#Neighborhood Data 
neighbourhood <- read_csv(paste(wd,"/data/Neighbourhood_POP.csv", sep=""))
```

```{r, include = FALSE}

# Convert data data type into character
chr <- as.character.Date(covidcases$`Episode Date`)
              
covidcases$'nd' <-chr 

# dates of interest

fweek <- c("2021-12-01","2021-12-02","2021-12-03","2021-12-04","2021-12-05","2021-12-06","2021-12-07")
lweek <- c("2021-12-25","2021-12-26","2021-12-27","2021-12-28","2021-12-29","2021-12-30","2021-12-31")

#filter for dates of interest 

fweektb <- subset(covidcases,nd %in% fweek)
lweektb <- subset(covidcases,nd %in% lweek)

ftab <- as.data.frame(table(fweektb$`Neighbourhood Name`)) 
ltab <- as.data.frame(table(lweektb$`Neighbourhood Name`)) 
```

```{r, include = FALSE}
covidcases$`Episode Date`[1]
```

```{r, include = FALSE}
#This section in creating final_table 

#Neighbourhood_Name, F_weekcases, L_weekcases, POP, F_weekrate,, L_weekrate, Rate_over_month, Case_over_month

#We need to add population for each neighbourhood and join it to the final table 

Neighbourhood <- ftab$Var1
F_weekcases <- ftab$Freq
L_weekcases <- ltab$Freq
Case_over_month <- ltab$Freq - ftab$Freq

#Create Final table 
Final_Table <- data.frame(Neighbourhood, F_weekcases, L_weekcases)

```

```{r, include = FALSE}
#3 Neighbourhood are different so there must be an manual correction prior to merging 

neighbourhood$Neighbourhood[128] <- "Weston-Pellam Park"
neighbourhood$Neighbourhood[16] <- "Briar Hill - Belgravia"
neighbourhood$Neighbourhood[30] <- "Danforth-East York"

```

```{r, include = FALSE}
#Merge neighbour with final table
Final_Table <- merge(Final_Table, neighbourhood)

# Creating the case rates per 10,000  
Final_Table$F_weekrate <- Final_Table$F_weekcases/Final_Table$POP * 10000
Final_Table$L_weekrate <- Final_Table$L_weekcases/Final_Table$POP * 10000

# Creating rate change 
Final_Table$Rate_Change <- Final_Table$L_weekrate - Final_Table$F_weekrate

# Creating fold increase 
Final_Table$fold_increase <- Final_Table$L_weekrate/Final_Table$F_weekrate 
```

```{r, include = FALSE}
#Neighbourhood Income Data
Socioeconomic_variables <- read_csv(paste(wd,"/data/Neighbourhoods_socioeconomic_variables.csv", sep=""))

```

```{r, include = FALSE}
#3 neighbourhoods are different so there must be an manual correction prior to merging 
Socioeconomic_variables$Neighbourhood[128] <- "Weston-Pellam Park"
Socioeconomic_variables$Neighbourhood[16] <- "Briar Hill - Belgravia"
Socioeconomic_variables$Neighbourhood[30] <- "Danforth-East York"
```

```{r, include = FALSE}
#Merge neighbourhood socioeconomic variables data with final table
Final_Table <- merge(Final_Table, Socioeconomic_variables)
```

```{r, include = FALSE}
#Merging shpfile to final table

nshp$FIELD_2[22] <- "Danforth-East York"
nshp$FIELD_2[25] <- "North St. James Town"
nshp$FIELD_2[93] <- "Mimico (includes Humber Bay Shores)"
nshp$FIELD_2[114] <- "Briar Hill - Belgravia"
nshp$FIELD_2[115] <- "Cabbagetown-South St. James Town"

nshp <- nshp %>%
  rename(Neighbourhood = FIELD_2)
Final_Table <- left_join(nshp, Final_Table, by = "Neighbourhood")

Final_Table <- Final_Table %>%
  rename(`_id` = FIELD_1,
          AREA_ATTR_ID = FIELD_3,
          PARENT_AREA_ID = FIELD_4,
          AREA_SHORT_CODE = FIELD_5,
          AREA_LONG_CODE = FIELD_6,
          AREA_NAME = FIELD_7,
          AREA_DESC = FIELD_8,
          X = FIELD_9,
          Y = FIELD_10,
          LONGITUDE = FIELD_11,
          LATITUDE = FIELD_12,
          OBJECTID = FIELD_13,
          Shape__Area = FIELD_14,
          Shape__Length = FIELD_15)

```

```{r, include = FALSE}
# to calculate population density
Final_Table$pop_density <- Final_Table$POP / Final_Table$Shape__Area
```


# Analysis

## Data Exploration

First, to obtain a preliminary understanding of trends in the data, choropleth maps were used to visualize the fold change in COVID-19 cases between the first and last week of December 2021 (Figure 2), per capita income (Figure 3), population density (Figure 4), and the proportion of 18- to 24-year-old individuals in each neighbourhood in Toronto (Figure 5).

```{r, echo = FALSE, fig.cap = "The change in COVID-19 rates (cases per 10,000) between the first and last week of December 2021 in Toronto by neighbourhood"}
# have change in COVID case rates (cases per 10000) between first and last week
# by difference and fold-change

ggplot(data = Final_Table) +
  geom_sf(aes(fill = fold_increase)) + 
  scale_fill_distiller(name = "Fold Change",
                       palette = "YlOrRd", 
                       direction = 1) +
  # ggtitle("Change in COVID-19 Rates (Cases per 10,000) Between the First and \n Last Week of December 2021 in Toronto by Neighbourhood") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

In Figure 2, it can be seen that the fold change for all neighbourhoods in Toronto is positive, indicating that they all experienced an increase in COVID-19 case rates in the last week of December 2021 compared to the first week of December 2021. In particular, the neighbourhoods coloured in a darker red and orange showed a significantly large fold change in COVID-19 rates in comparison to other neighbourhoods, which possibly signifies the location of COVID-19 hotspots during December 2021. Accordingly, the neighbourhoods located close to such spots seem to have relatively large fold changes as well, as shown through their light orange colour. Neighbourhoods with smaller fold changes, seen in yellow in the map, appear to be located further away from areas with more intense fold changes.

```{r, include = FALSE}
# rename column names so it's easier to call
Final_Table <- Final_Table %>%
  rename(avg_income = `Average income $ (Total income in 2010 of population aged 15 years and over)`,
         perc_18_24 = `Percent of Pop Age 18 to 24`)
```

```{r, echo = FALSE, fig.cap = "The average income of individuals in Toronto by neighbourhood"}
ggplot(data = Final_Table) +
  geom_sf(aes(fill = avg_income)) + 
  scale_fill_distiller(name = "Average Income ($)",
                       palette = "YlOrRd", 
                       direction = 1) +
  # ggtitle("Average Income in Toronto by Neighbourhood") #maybe adjust this later for detail +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

Neighbourhoods with greater average income appear darker in colour on the map in Figure 3, showing that people with higher income tend to concentrate near the center of Toronto. Other neighbourhoods with high average income can be seen in the west of Toronto as well. 

From a visual comparison of Figures 2 and 3, it looks like the neighbourhoods with higher average incomes have lower fold changes in COVID-19 case rates while the neighbourhoods with high fold changes are those with lower average incomes. 


```{r, echo = FALSE, fig.cap = The population density of each of Toronto’s neighbourhoods}

ggplot(data = Final_Table) +
  geom_sf(aes(fill = pop_density)) + 
  scale_fill_distiller(name = "People per sqaure km",
                       palette = "YlOrRd", 
                       direction = 1) +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
```

From Figure 4, it seems that there may be a greater population density in neighbourhoods located towards the southern and northeastern areas of Toronto. Population density appears to be lower in neighbourhoods located in the outskirts of the city.
[insert choropleth map for population prop]

```{r, echo = FALSE, fig.cap = "The proportion of 18- to 24-year-old individuals in Toronto by neighbourhood"}
ggplot(data = Final_Table) +
  geom_sf(aes(fill = perc_18_24)) + 
  scale_fill_distiller(name = "Population Percentage",
                       palette = "YlOrRd", 
                       direction = 1) +
  # ggtitle("Proportion of People Aged 18-24 in Toronto by Neighbourhood") #maybe adjust this later for detail (e.g., as of...) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

Looking at Figure 5, there are a cluster of neighbourhoods in the south of Toronto, coloured in red, that have a particularly high percentage of 18- to 24-year-old individuals making up their population. In addition, a neighbourhood located near the north, in dark orange, has a relatively high proportion of 18- to 24-year-old individuals. 

## Spatial autocorrelation
Before doing a regression analysis to determine the relationship between the change in COVID-19 case rates and the predictor variables, an analysis of the pattern in the change in COVID-19 case rates must first be done. If its pattern is non-random, then regression analysis can be used to provide more information about the process that is creating this pattern.

To test whether a spatial pattern is random or non-random, Moran’s I, a coefficient of spatial autocorrelation, can be calculated and used in a hypothesis test that assesses for spatial randomness. The null hypothesis of this test is that the pattern is spatially random. The results of the Moran’s I test on the change in COVID-19 case rates can be seen in Figure 6.

```{r, echo = FALSE, fig.cap = "The results of the hypothesis test using Moran’s I on the change in COVID-19 case rates between the first and last week of December 2021 in Toronto."}

moran.test(Final_Table$fold_increase, Final_Table.w)
```

From Figure 6, the Moran’s I statistic for this data is 0.089308782 and the p-value of the test is 0.01787. Since the p-value is less than 0.05, it is sufficiently small enough to reject the null hypothesis with a high degree of confidence. This means that the change in COVID-19 case rates is spatially autocorrelated, not spatially random. As well, since the Moran’s I statistic is positive, it suggests that it is a non-random spatial pattern where high values tend to be with other high values, and low values tend to be with other low values. This agrees with the visual observation of the choropleth map of the change in COVID-19 case rates in Figure 2. 

# Conclusion

This study shows that there is a clear connection between a lower-income status and the incidence rates of the COVID-19. The results show that other socio-economic factors, such as age or population density did not have a clear connection with the data currently available. Socio-economic factors are therefore complex factors that need to be further explored. Identifying these groups from a socio-economic perspective is the first step to establishing that a person’s SEP may be as much of an indicator in predicting health outcomes as a pre-existing medical condition. In establishing such similarities, governing healthcare bodies may extend precautionary recommendations to people with specific socio-economic conditions, thus providing a more in-depth and informed disease prevention plan. In a pursuit of identifying groups of the populations who are more susceptible to poorer outcomes when combating health problems, more research needs to be undertaken. 

# References
