---
title: "Income as a Determinant of the Speed of Transmission: Study of the Thrid Wave of SARS-CoV-2 in Toronto, Ontario"
#subtitle: "A subtitle"
thanks: ""
author:
- name: Jason Lam
  student_number: 400201299
- name: 
  student_number: 
- name: 
  student_number: 
- name: 
  student_number: 
- name: 
  student_number: 
subject: "ENVSOCTY 4GA3"
abstract: "This paper reports our analysis of opioid prescriptions in Kentucky, USA, and its relationship with income and education at the level of counties. Data were obtained from the Washington Post database and the US Census."
keywords: "opioids, education, income, spatial analysis"
date: "4/22/2022"
output:
  pdf_document:
    # The project-template-default.tex file was minimally adapted from Steven V. Miller's template for academic manuscripts. See:
    # http://svmiller.com/blog/2016/02/svm-r-markdown-manuscript/
    # https://github.com/svmiller/svm-r-markdown-templates/blob/master/svm-latex-ms.tex
    template: project-template-default.tex
bibliography: [bibliography.bib, packages.bib]
always_allow_html: true
---

<!--Chunks of code can have names; the chunk option "include" controls whether the chunk and its output are printed in the final document-->
```{r load-packages, include=FALSE}

```

```{r write-package-bib, include=FALSE}
# This function is used to write a bibliography for the `R` packages used in the paper
knitr::write_bib(file = 'packages.bib')
```

```{r read-census-data, include=FALSE}
```

<!--This create a page break, i.e., starts a new page-->
<!--\newpage-->

<!-- 
To cite references in your bibliography.bib file, use [@item] if you want it to be cited in brackets, or @item if you want it to be cited as Name (year). If you want to cite various items in brackets, separate them with semicolons [@item1; @item2]
-->


# Introduction

- Describe the when COVID first cases in Toronto, 
- First and Second wave, policy the government implemented 
- What lead to the third wave 
  - Relaxing of restrictions 
  - Emergence of Omnicron 

# Study area

*varaible* and COVID-19 incident was examined at the neighbourhood level for the City of Toronto. In total there are 140 unique areas. There were created by City of Toronto to help government and other planning oganizations with obtaining socio-economic data. The boundaries are based around the Canadian Census Tracks. Each niehgbourhood may contain between two to five census tracts. [@Toronto_2017]

# Data


# Methods


# Results


# Analysis


# Conclusion


# References
